---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

This document outlines the steps taken to read and process a set of 
activity monitoring data.

*Read In Data:*

```{r readin}
        actData<-read.csv("activity.csv")
        
        ##coerce dates to date type
        actData$date<-as.Date(actData$date,"%Y-%m-%d")

```
Once the data has been read in we can undertake initial analysis of the data:
```{r histday}
                library(ggplot2)
        
        plotObj<-ggplot(actData,aes(date,steps)) + 
                stat_summary(fun.y=sum, geom="bar") + labs(title
                ="Steps by Day")
        
        print(plotObj)
```

Mean & Median Steps per day calculated as follows
```{r mean_med,results='asis'}
   library(xtable)
        
       daymean<-tapply(actData$steps,actData$date,mean, na.rm=TRUE)
       daymedian<-tapply(actData$steps,actData$date,median, na.rm=TRUE)
       
       df<-data.frame(mean_steps=daymean,median_steps=daymedian)
       xt<-xtable(df)
       print(xt, type="html")
             
```    

Reviewing the number of steps per interval we have created the following plot
```{r steps_int}
        plotObj<-ggplot(actData,aes(interval,steps)) + 
                stat_summary(fun.y=mean,geom="line") + labs(title
                                                          ="Steps by Day")
        
        print(plotObj)
        avestep<-tapply(actData$steps,actData$interval,mean,na.rm=TRUE)
        maxInt<-max(avestep)
        avestep[avestep==maxInt]
```
The maximum average number of steps in an interval is `r maxInt` in interval 835

```{r no_NAs}
   num_NAs<-sum(is.na(actData$steps))
             
```

The original dataset contains `r num_NAs` NA values.

## Imputing missing values
I've decided to use the following strategy where the number of steps is missing
- Where the mean number of steps per interval is available use this
- otherwise use 0 - given a large number of these are liekly to be overnight
with limited movement

```{r impute}
        library(plyr)
        missingvals<-is.na(actData$steps)
        no_missing<-nrow(actData[missingvals,])
        
        #generate lookup sequence
        x<-seq(0,2355,5)
        y<-rep(0,length(x))
        df<-data.frame(interval=x,avesteps=y)
        df2<-data.frame(interval=rownames(avestep),as=avestep)
        df2$interval<-as.character(df2$interval)
        df2$interval<-as.numeric(df2$interval)
        df3<-join(df,df2)
        df3$calcstep<-ifelse(is.na(df3$as),0,df3$as)
        
        actData2<-actData
        actData2$date<-as.Date(actData2$date, "%Y-%m-%d")
        actData2red<-actData[missingvals,]
        actData2red$date<-as.Date(actData2red$date, "%Y-%m-%d")
        
        for (i in 1:no_missing) {
                #get interval
                reqint<-actData2red$interval[i]
                
                reqdate<-actData2red$date[i]
                
                #returnval
                        actData2$steps[actData2$interval==reqint & actData2$date==reqdate]<-df3$calcstep[df3$interval==reqint]
                }
        
        newNAs<-sum(is.na(actData2$steps))     
```

The number of NAs now present in the data is `r newNAs`

The new data set now looks as follows:

```{r histday2}
                
        plotObj<-ggplot(actData2,aes(date,steps)) + 
                stat_summary(fun.y=sum,geom="bar") + labs(title
                ="Steps by Day")
        
        print(plotObj)
```

The comparison to the previous mean and media values is shown in the resultant table

```{r mean_med2,results='asis'}
   library(xtable)
        
       daymean<-tapply(actData$steps,actData$date,mean, na.rm=TRUE)
       newmean<-tapply(actData2$steps,actData2$date,mean, na.rm=TRUE)
       mean.dif<-newmean-daymean
       daymedian<-tapply(actData$steps,actData$date,median, na.rm=TRUE)
       newmedian<-tapply(actData2$steps,actData2$date,median, na.rm=TRUE)
       
       
       df4<-data.frame(mean_steps=daymean,new_mean=newmean,diff=mean.dif,
                      median_steps=daymedian,new_median=newmedian)
       xt<-xtable(df4)
       print(xt, type="html")
             
``` 

### Is there any difference between Weekend & Weekday activity profiles?

First we need to create a factor indicating that the day is a weekend or week day:

```{r weekday_factor}

       actData2$wd<-weekdays(actData2$date)
       wdclass<-grep("Saturday|Sunday",actData2$wd)
       #wdclass
       actData2$weekdayclass<-NA
       actData2$weekdayclass[wdclass]<-"Weekend"
       actData2$weekdayclass[is.na(actData2$weekdayclass)]<-"Weekday"
        actData2$weekdayclass<-as.factor(actData2$weekdayclass)
        #str(actData2)


```

Now create a panel plot to display the data

```{r weekday panel}
       plotObj<-ggplot(actData2,aes(interval,steps)) + 
                stat_summary(fun.y=mean,geom="line") + labs(title
                ="Steps by Day") + facet_grid(weekdayclass~.)
        
        print(plotObj) 
        
                
```

The plot shows that typically activity egins earlier and with a greater burst than
at the weekend  